---
title: "ENSAB"
author: "Maria Camila Restrepo"
date: "2025-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ENSAB databases

```{r}
library(here)
library(readr)
library(tidyverse)
if(!require(MASS)) install.packages("MASS")
library(MASS)
if(!require(survey)) install.packages("survey")
library(survey)

ninoscompleta <- read_delim(
  here("ensab", "M3_NINOS.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

examen13completa <- read_delim(
  here("ensab", "M4_EXAMEN_1.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

examen5completa <- read_delim(
  here("ensab", "M4_EXAMEN_2.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

hogarcompleta <- read_delim(
  here("ensab", "M1_HOGAR.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

```

We have a total of 6446 children for the sample size

```{r}
glimpse(ninoscompleta)
names(ninoscompleta)
```

Keep variables:

- MPIO
- DPTO
- M3_100C: age
- M3_113D: grupo etnico
- M3_113F: maximo nivel educativo alcanzado por el ppal responsable economico
- M3_119: ingresos mensuales ppal responsable economico
- M3_201: Where does the child spend most of their time
- M3_202: With whom does the child spend most of their time
- M3_224: Lactancia materna edad
- M3_225: Biberon
- M3_226: Biberon edad
- M3_227: Que le dan en el biberon. 227A-227CU
- Region: a donde pertenece el municipio
- Subreg:
- Municipio:
- Estrato: Estrato muestral dentro de subregion
- Fex_fin: factor de expansion final
- Sexo
- Zona: urbana, rural
- EST_SE_2: Nivel socioeconomico
- FRACCION: fraccion social
- REGIMEN: contributivo, subsidiado, especial, no asegurado
- ID_HOGAR
- ID_PERSONA

Grupo etareo was excluded because it is the same as age.

```{r}
ninos <- ninoscompleta %>% 
  dplyr::select(ÿDANE, MPIO, DPTO, M3_100C, M3_113D, M3_113E, M3_113F, M3_119, M3_201, M3_202, M3_224, M3_225, M3_226, M3_227, M3_227A, M3_227B, M3_227C, M3_227D, M3_227E, M3_227F, M3_227G, M3_227CU, Region, Subreg, Municipio, Estrato, Fex_fin, Sexo, Zona, EST_SE_2, FRACCION, REGIMEN, ID_HOGAR, ID_PERSONA)

glimpse(ninos)

ninoscompleta %>%
  group_by(MPIO) %>%
  summarise(unique_DANE = n_distinct(ÿDANE)) %>%
  arrange(desc(unique_DANE))

#We can see there are some municipalities by the same name, so it is better to use the unique identifier from the census. 

ninos <- ninos %>%
  rename(census_mpio = ÿDANE)
```

Hogar has the same basic variables, so we wont join it for now.

Exploring hogar:

```{r}
#Ethnicity

hogarcompleta %>%
  group_by(M1_303D) %>%
  count()

#14238 NAs. Won't use it
```


## Main outcome: dmft index

Caries:

DMFT: decayed, missing and filled teeth

Questions 305 and 307

0: A Sano, F sellante, 
1: B Cavitaded, C: filling with caries, D: filled without caries, E: missing due to caries, G: Protesis, corona, retenedor, carilla
H: no registra (ausente o sin erupcionar)

MA_305A, MA_305B, MA_305C, MA_305D, MA_305E, MA_305F, MA_305G, MA_305H, MA_305I, MA_305J
MA_307A, MA_307B, MA_307C, MA_307D, MA_307E, MA_307F, MA_307G, MA_307H, MA_307I, MA_307J

For permanent teeth: 

0: 0 Sano, 5: missing for another reason, 6: seallant, 8: not erupted, 9: not registered (ausente)
1: 1 cavitated, 2: filling with caries, 3: filled without caries, 4: missing due to caries, 7: Protesis, corona, retenedor, carilla

MB_208: Aplicación del examen clínico (dependiendo de 
antecedentes médicos). 1: se aplica el examen. 2: no se aplica el examen.

### Children 1 and 3 years old

```{r}
examen13 <- examen13completa %>% 
  dplyr::select(MA_201B, MA_208, MA_305A, MA_305B, MA_305C, MA_305D, MA_305E, MA_305F, MA_305G, MA_305H, MA_305I, MA_305J, MA_307A, MA_307B, MA_307C, MA_307D, MA_307E, MA_307F, MA_307G, MA_307H, MA_307I, MA_307J, ID_HOGAR, ID_PERSONA)

#Assign a value to 1 to the following letters: B, C, D, E, G
examen13 <- examen13 %>%
  mutate(
    across(
      starts_with("MA_3"),
      ~case_when(
        str_detect(., "[BCDEG]") ~ 1,
        TRUE ~ 0)
    ))

#create variable dmft
examen13 <- examen13 %>%
  mutate(dmft = rowSums(across(starts_with("MA_3"), ~. > 0))) %>%
  dplyr::select(-starts_with("MA_3"))

glimpse(examen13)
summary(examen13$dmft)

#See distribution
ggplot(data = examen13, aes(x = dmft)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(x = "dmft Index score", y = "Participants", title = "Distribution of dmft Index score")
```

### Children 5 years old

dmft can take into account the first permanent molar. So the maximum value can be up to 24.

```{r}
examen5 <- examen5completa %>% 
  dplyr::select(MB_201B, MB_208, MB_305A, MB_305B, MB_305C, MB_305D, MB_305E, MB_305F, MB_305G, MB_305H, MB_305I, MB_305J, MB_305K, MB_305L, MB_307A, MB_307B, MB_307C, MB_307D, MB_307E, MB_307F, MB_307G, MB_307H, MB_307I, MB_307J, MB_307K, MB_307L, ID_HOGAR, ID_PERSONA)

glimpse(examen5)

#Assign a value to 1 to the following numbers: 1, 2, 3, 4, 7
examen5 <- examen5 %>%
  mutate(
    across(
      starts_with("MB_3"),
      ~case_when(
        str_detect(., "^[12347]$|^[BCDEG]$") ~ 1,
        TRUE ~ 0                                 
      )))

#create variable dmft
examen5 <- examen5 %>%
  mutate(dmft = rowSums(across(starts_with("MB_3"), ~. > 0))) %>%
  dplyr::select(-starts_with("MB_3"))

summary(examen5$dmft)

#See distribution
ggplot(data = examen5, aes(x = dmft)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(x = "dmft Index score", y = "Participants", title = "Distribution of dmft Index score")
```

Merge datasets:

```{r}
ensab <- ninos %>%
  left_join(examen13, by = "ID_PERSONA") %>%
  left_join(examen5, by = "ID_PERSONA") %>%
  mutate(dmft = coalesce(dmft.x, dmft.y), 
         ID_HOGAR = coalesce(ID_HOGAR.x, ID_HOGAR.y)) %>%
  dplyr::select(-ID_HOGAR.x, -ID_HOGAR.y, -dmft.x, -dmft.y)

#Missing data

ensab %>%
  filter(MA_208 == 2 | MB_208 == 2) %>%
  count()

ensab <- ensab %>%
  filter(MA_208 == 1 | MB_208 == 1) %>%
  dplyr::select(-MB_208, -MA_208)

#only  3 children did not have the clinical examination portion. We'll exclude them. Now the sample size is 6443 children.
```

## Account for survey design and weights

Primary sampling unit: municipality

Sampling strata: subregion, named as Estrato in the data set. 

```{r}
survensab <- svydesign(ids = ~census_mpio, data = ensab, strata = ~Estrato, weights = ~Fex_fin, nest = TRUE)
summary(survensab)
glimpse(ensab)

#checking how many children share the same household

ensab %>%
  group_by(ID_HOGAR) %>%
  count() %>%
  arrange(desc(n)) %>%
  filter(n > 1) %>%
  rename(cph = n) %>%
  group_by(cph) %>%
  summarise(prop = n() / nrow(ensab) * 100)

#737 children share the same household with another child included in the study. We won't take it into account for now.

```


#Key variables

Age: M3_100C

It really is categorical and not continuous.

```{r}
glimpse(ensab$M3_100C)
summary(ensab$M3_100C)

#rename M2_100C to age
ensab <- ensab %>%
  rename(age = M3_100C) %>%
  mutate(age = factor(age, 
                      levels = c(1, 3, 5),
                      labels = c("1", "3", "5")))

summary(ensab$age)
```


## Exposure

### Breastfeeding age in months

M3_224

91: Never
92: Doesn't know (NA)

```{r}
glimpse(ensab$M3_224)

ensab %>%
  filter(M3_224 == 92) %>%
  count()
#98 NAs

ensab <- ensab %>%
  mutate(bf = case_when(
    M3_224 == 92 ~ NA,
    TRUE ~ M3_224))

#distribution of bf

ensab %>%
  filter(!(bf == 91)) %>%
  ggplot(aes(x = bf)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(ensab$bf, na.rm = TRUE), max(ensab$bf, na.rm = TRUE), by = 6)) +
  labs(x = "Breastfeeding age", y = "Participants", title = "Distribution of breastfeeding age")

ensab %>%
  filter(!(bf == 91)) %>%
  summary()

#12 seems like the maximum value

#Categorize into: never, 0-6 months, more than 6 months

ensab <- ensab %>%
  mutate(bf = case_when(
    bf == 91 ~ 1,
    bf <= 6 ~ 2,
    bf > 6 ~ 3,
  )) %>%
  mutate(bf = factor(bf, 
                     levels = c(1, 2, 3),
                     labels = c("Never", "0-6 months", "More than 6 months")))

summary(ensab$bf)
         
```

### Bottle use: M3_225, M3_226, M3_227, M3_227A-M3_227CU, and M3_228

For M3_227, we want to include:
- D: sugared drinks
- F: sodas
- G: others, explore which ones and if they are cariogenic or not
Exclude A (mothers milk), B (formula), C (other types of milk).
E is unsure if it includes cariogenic drinks

```{r}
#Bottle use at the moment of the survey

ensab %>%
  filter(M3_225 == 2) %>%
  count()
#4487 don't use the bottle currently

#proportion of those who use and dont use bottle: variable M3_225

prop.table(table(ensab$M3_225)) #70% don't use the bottle

glimpse(ensab$age)

#of those that do, what age are they?
ensab %>%
  filter(M3_225 == 1) %>%
  group_by(age) %>%
  count()

#225 and 226 include past and current use, might be a problem to create variable of age of bottle use. It might be more important what is in the bottle.

#M3_226:

ensab %>%
  filter(M3_226 == 91) %>%
  count()
#1879 never were given bottle

1879/6443*100

ensab %>%
  filter(M3_226 == 92) %>%
  count()
#30 NAs


#M3_227: what do they put in the bottle. 0: no aplica, 1: aplica

glimpse(ensab$M3_227CU)

#ensab %>%
#  filter(M3_227G == 1) %>%
#  select(M3_227G, M3_227CU, ID_PERSONA) %>% #order M3_227CU in alphabetical order
#  distinct(M3_227CU)
  
```

## Possible confounders or EMM

### Demographic variables

Sex: Sexo

Couldn't find the codification in the codebook so I had to guess.

```{r}
glimpse(ensab$Sexo)

ensab <- ensab %>%
  mutate(sex = factor(Sexo,
                         levels = c(1, 2),
                         labels = c("Masculine", "Feminine")))

summary(ensab$sex)
```

Ethnicity: M3_113D

Decided not to include ethnicity because of 5772 missing values

```{r}
glimpse(ensab$M3_113D)

ensab %>%
  group_by(M3_113D) %>%
  count()
```

### Socioeconomic position

Socioeconomic status: EST_SE_2
```{r}
glimpse(ensab$EST_SE_2)

ensab <- ensab %>%
  mutate(ses = factor(EST_SE_2, 
                           levels = c(1, 2, 3, 4), 
                           labels = c("Low low", "Low", "Middle low", "Middle to high")))

summary(ensab$ses)
```

Income: M3_119

NMW: national minimum wage (salario minimo legal vigente)

```{r}
glimpse(ensab$M3_119)

ensab %>%
  group_by(M3_119) %>%
  count()

#8 are NAs
ensab %>%
  filter(M3_119 == 8) %>%
  count() #95 NAs

#transform 7 unemployed into 0

ensab <- ensab %>%
  mutate(income = case_when(
    M3_119 == 8 ~ NA,
    M3_119 == 7 ~ 0,
    TRUE ~ M3_119)) %>%
  mutate(income = factor(income,
                         levels = c(0, 1, 2, 3, 4, 5, 6),
                         labels = c("Unemployed", "Less than half NMW", "Half - 1 NMW", "1-2 NMW", "2-3 NMW", "3-4 NMW", "More than 4 NMW")))

summary(ensab$income)

```

Educational level of the provider: M3_113F

Decided to exclude because there are 5772 missing values

```{r}
glimpse(ensab$M3_113F)

ensab %>%
  group_by(M3_113F) %>%
  count()

```


Health insurance coverage/scheme:

Chose REGIMEN over M3_113E because it has less categories plus no missing values, and it reflects the insurance status of the child and not the main provider.

```{r}
glimpse(ensab$REGIMEN)

#counting NAs in M3_113E

ensab %>%
  filter(M3_113E == 6) %>%
  count()

ensab <- ensab %>%
  mutate(insurance = case_when(
    REGIMEN == 4 ~ 0,
    TRUE ~ REGIMEN)) %>%
  mutate(insurance = factor(insurance, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Uninsured", "Contributory", "Subsidized", "Other")))

summary(ensab$insurance)
```

Region

```{r}
glimpse(ensab$Region)

ensab <- ensab %>%
  mutate(region = factor(Region,
                         levels = c(1, 2, 3, 4, 5, 6),
                         labels = c("Atlantic", "Eastern", "Central", "Pacific", "Bogota", "Orinoquia - Amazonia")))

summary(ensab$region)
```

Urban vs rural: Zona

Will be named location
Chose Zona over Clase because it has more universal categories (rural and urban), while Clase has categories according to the national census. However, it has more granularity and could be useful.

```{r}
glimpse(ensab$Zona)

ensab %>%
  group_by(Zona, .drop = FALSE) %>%
  count()

ensab <- ensab %>%
  mutate(location = factor(Zona,
                           levels = c(1, 2),
                           labels = c("Urban", "Rural")))

summary(ensab$location)
```

### Potentially interesting variables to check for EMM

M3_201: Where the child spends most of their time
Could categorize into home or not home

M3_202: Who takes care of the child
1: Mother or father
2: Relative who is overage

```{r}
ensab %>%
  group_by(M3_201) %>%
  count()

ensab %>%
  group_by(M3_202) %>%
  count()
```


# Association between caries and possible exposures

## Exploring the relationship between age of first appointment and caries (dmft)

The hypothesis is that the older the child was at it's first appointment, the higher the dmft index.

ensab %>%
  filter(!is.na(agefdv)) %>%
ggplot(aes(x=agefdv, y=dmft)) +
  geom_boxplot() +
  facet_wrap(~age) +
  labs(x = "Age of first dental visit", y = "dmft Index score", title = "dmft Index score by age of first dental visit")

```{r}

```

