---
title: "ENSAB"
author: "Maria Camila Restrepo"
date: "2024-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ENSAB databases

```{r}
library(here)
library(readr)
#install package MASS if not installed previously
if(!require(MASS)) install.packages("MASS")
library(MASS)
library(tidyverse)

ninoscompleta <- read_delim(
  here("ensab", "M3_NINOS.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

examen13completa <- read_delim(
  here("ensab", "M4_EXAMEN_1.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

examen5completa <- read_delim(
  here("ensab", "M4_EXAMEN_2.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

hogarcompleta <- read_delim(
  here("ensab", "M1_HOGAR.txt"), 
  delim = "|", 
  locale = locale(encoding = "Latin1")
)

```

We have a total of 6446 children for the sample size

```{r}
glimpse(ninoscompleta)
names(ninoscompleta)
```

Keep variables:

- MPIO
- DPTO
- M3_100C: age
- M3_113E: regimen del ppal responsable economico
- M3_113F: maximo nivel educativo alcanzado por el ppal responsable economico
- M3_119: ingresos mensuales ppal responsable economico
- M3_205: ha tenido cita con odontologo
- M3_206A: Edad de primera cita
- M3_206B: Meses o años. 1: meses, 2 años.
- M3_208: Lugar de consulta
- M3_209: Motivo de connsulta
- M3_210: No atencion
- M3_213: Edad de comenzar higiene
- M3_214: Quien hace higiene
- M3_224: Lactancia materna edad
- M3_218: Informacion sobre tipo de crema dental
- M3_225: Biberon
- M3_226: Biberon edad
- M3_227: Que le dan en el biberon. 227A-227CU
- Clase: cabecera, cp, rural
- Region: a donde pertenece el municipio
- Subreg:
- Municipio:
- Estrato: Estrato muestral dentro de subregion
- Fex_fin: factor de expansion final
- Sexo
- Zona: urbana, rural
- EST_SE_2: Nivel socioeconomico
- FRACCION: fraccion social
- REGIMEN: contributivo, subsidiado, especial, no asegurado
- ID_HOGAR
- ID_PERSONA

Grupo etareo was excluded because it is the same as age.

Hogar has the same basic variables, so we wont join it for now.

```{r}
ninos <- ninoscompleta %>% 
  dplyr::select(MPIO, DPTO, M3_100C, M3_113E, M3_113F, M3_119, M3_205, M3_206A, M3_206B, M3_207, M3_208, M3_209, M3_210, M3_213, M3_214, M3_218, M3_224, M3_225, M3_226, M3_227, M3_227A, M3_227B, M3_227C, M3_227D, M3_227E, M3_227F, M3_227G, M3_227CU, Clase, Region, Subreg, Estrato, Fex_fin, Sexo, Zona, EST_SE_2, FRACCION, REGIMEN, ID_HOGAR, ID_PERSONA)

glimpse(ninos)
```

## Main outcome: dmft index

Caries:

DMFT: decayed, missing and filled teeth

Questions 305 and 307

0: A Sano, F sellante, 
1: B Cavitaded, C: filling with caries, D: filled without caries, E: missing due to caries, G: Protesis, corona, retenedor, carilla
H: no registra (ausente o sin erupcionar)

MA_305A, MA_305B, MA_305C, MA_305D, MA_305E, MA_305F, MA_305G, MA_305H, MA_305I, MA_305J
MA_307A, MA_307B, MA_307C, MA_307D, MA_307E, MA_307F, MA_307G, MA_307H, MA_307I, MA_307J

For permanent teeth: 

0: 0 Sano, 5: missing for another reason, 6: seallant, 8: not erupted, 9: not registered (ausente)
1: 1 cavitated, 2: filling with caries, 3: filled without caries, 4: missing due to caries, 7: Protesis, corona, retenedor, carilla

MB_208: Aplicación del examen clínico (dependiendo de 
antecedentes médicos). 1: se aplica el examen. 2: no se aplica el examen.

### Children 1 and 3 years old

```{r}
examen13 <- examen13completa %>% 
  dplyr::select(MA_208, MA_305A, MA_305B, MA_305C, MA_305D, MA_305E, MA_305F, MA_305G, MA_305H, MA_305I, MA_305J, MA_307A, MA_307B, MA_307C, MA_307D, MA_307E, MA_307F, MA_307G, MA_307H, MA_307I, MA_307J, ID_HOGAR, ID_PERSONA)

#Assign a value to 1 to the following letters: B, C, D, E, G
examen13 <- examen13 %>%
  mutate(
    across(
      starts_with("MA_3"),
      ~case_when(
        str_detect(., "[BCDEG]") ~ 1,
        TRUE ~ 0)
    ))

#create variable dmft
examen13 <- examen13 %>%
  mutate(dmft = rowSums(across(starts_with("MA_3"), ~. > 0))) %>%
  dplyr::select(-starts_with("MA_3"))

glimpse(examen13)
summary(examen13$dmft)

#See distribution
ggplot(data = examen13, aes(x = dmft)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(x = "dmft Index score", y = "Participants", title = "Distribution of dmft Index score")
```

### Children 5 years old

dmft can take into account the first permanent molar. So the maximum value can be up to 24.

```{r}
examen5 <- examen5completa %>% 
  dplyr::select(MB_208, MB_305A, MB_305B, MB_305C, MB_305D, MB_305E, MB_305F, MB_305G, MB_305H, MB_305I, MB_305J, MB_305K, MB_305L, MB_307A, MB_307B, MB_307C, MB_307D, MB_307E, MB_307F, MB_307G, MB_307H, MB_307I, MB_307J, MB_307K, MB_307L, ID_HOGAR, ID_PERSONA)

glimpse(examen5)

#Assign a value to 1 to the following numbers: 1, 2, 3, 4, 7
examen5 <- examen5 %>%
  mutate(
    across(
      starts_with("MB_3"),
      ~case_when(
        str_detect(., "^[12347]$|^[BCDEG]$") ~ 1,
        TRUE ~ 0                                 
      )))

#create variable dmft
examen5 <- examen5 %>%
  mutate(dmft = rowSums(across(starts_with("MB_3"), ~. > 0))) %>%
  dplyr::select(-starts_with("MB_3"))

summary(examen5$dmft)

#See distribution
ggplot(data = examen5, aes(x = dmft)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(x = "dmft Index score", y = "Participants", title = "Distribution of dmft Index score")
```

Merge datasets:

```{r}
ensab <- ninos %>%
  left_join(examen13, by = "ID_PERSONA") %>%
  left_join(examen5, by = "ID_PERSONA") %>%
  mutate(dmft = coalesce(dmft.x, dmft.y), 
         ID_HOGAR = coalesce(ID_HOGAR.x, ID_HOGAR.y)) %>%
  dplyr::select(-ID_HOGAR.x, -ID_HOGAR.y, -dmft.x, -dmft.y)

#Missing data

ensab %>%
  filter(MA_208 == 2 | MB_208 == 2) %>%
  count()

ensab <- ensab %>%
  filter(MA_208 == 1 | MB_208 == 1) %>%
  dplyr::select(-MB_208, -MA_208)

#only  3 children did not have the clinical examination portion. We'll exclude them. Now the sample size is 6443 children.
```

# Posible exposures

Age of first dental visit (agefdv). It is composed of:
- M3_206A: numeric
- M3_206B: 1: months, 2: years

Take into account M3_205: have they ever had an appointment with a dentist?  
1: yes, 2: no, 3: doesn't know

Combine 206A and 205 to categorize.

```{r}

#Turn M3_206A into a numeric variable
ensab$M3_206A <- as.numeric(ensab$M3_206A)

summary(ensab$M3_206A)

ensab %>%
  filter(M3_205 == 2 | M3_205 == 3) %>%
  count()
#2648: coincides with the number of NAs

ensab %>%
  filter(M3_205 == 3) %>%
  count()
#only 46 participants have missing values for M3_205

#multiply values in the filtered rows by 12 and categorize
ensab <- ensab %>%
  mutate(agefdv = if_else(M3_206B == 2, M3_206A * 12, M3_206A))

ggplot(ensab, aes(x = agefdv)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(ensab$agefdv, na.rm = TRUE), max(ensab$agefdv, na.rm = TRUE), by = 6)) +
  labs(x = "Age of first dental visit", y = "Participants", title = "Distribution of age of first dental visit")

# can do bins 0-6, 7-12, 13-24, 25-36, 37-48, 49-72, or just divide into 3.

ensab <- ensab %>%
  mutate(agefdv = case_when(
    M3_205 == 3 ~ NA_real_,
    M3_205 == 2 ~ 0,
    agefdv <= 12 ~ 1,
    agefdv > 12 ~ 2,
    TRUE ~ agefdv)) %>%
  mutate(agefdv = factor(agefdv, 
                         levels = c(0, 1, 2), 
                         labels = c("Never", "0-12", "13+")))

summary(ensab$agefdv)

ensab <- ensab %>%
  dplyr::select(-M3_206B, -M3_206A)
```

Reason for last consultation (reason)

M3_209: 
1: por una urgencia
2: para un tratamiento
3: por revision/prevencion
4: certificado odontologico

M207 Is important: when was appointment of the child with the dentist?
1: less than a year ago
2: between 1-2 years
3: more than 3 years
4: does not remember (NA)

```{r}
ensab %>%
  filter(M3_207 == 4) %>%
  count()

#45 NAs to be expected in M3_209

ensab <- ensab %>%
  mutate(reason = factor(M3_209, 
    levels = c(1, 2, 3, 4), 
    labels = c("Emergency", "Treatment", "Prevention", "Certification")))

summary(ensab$reason)
#2648 NAs
```

Age of start of dental hygiene

M3_213: units are months, continuous
91: doesn't know
92: has never gotten dental hygiene done

```{r}
ensab %>%
  filter(M3_213 == 91 | M3_213 == 92) %>%
  count()
#385

glimpse(ensab$M3_213)

#create distribution
ensab %>%
  filter(!(M3_213 %in% c(91, 92))) %>%
  ggplot(aes(x = M3_213)) +
  geom_histogram(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = seq(min(ensab$M3_213, na.rm = TRUE), max(ensab$M3_213, na.rm = TRUE), by = 6)) +
  labs(x = "Age of start of dental hygiene", y = "Participants", title = "Distribution of age of start of dental hygiene")
```

Information on toothpaste for the child:

M3_219

```{r}
glimpse(ensab$M3_218)

ensab <- ensab %>%
  mutate(tp_info = case_when(
    M3_218 == 3 ~ NA,
    TRUE ~ M3_218)) %>%
  mutate(tp_info = factor(tp_info,
                          levels = c(1, 2),
                          labels = c("Yes", "No")))

summary(ensab$tp_info)
```

```{r}
# who does the dental hygiene

# breastfeeding age

# bottle
```

# Possible confounders or EMM

Age: M3_100C

It really is categorical and not continuous

```{r}
glimpse(ensab$M3_100C)
summary(ensab$M3_100C)

#rename M2_100C to age
ensab <- ensab %>%
  rename(age = M3_100C) %>%
  mutate(age = factor(age, 
                      levels = c(1, 3, 5),
                      labels = c("1", "3", "5")))

summary(ensab$age)
```

Sex: Sexo

Couldn't find the codification in the codebook so I had to guess

```{r}
glimpse(ensab$Sexo)

ensab <- ensab %>%
  mutate(sex = factor(Sexo,
                         levels = c(1, 2),
                         labels = c("Masculine", "Feminine")))

summary(ensab$sex)
```

Socioeconomic status: EST_SE_2
```{r}
glimpse(ensab$EST_SE_2)

ensab <- ensab %>%
  mutate(ses = factor(EST_SE_2, 
                           levels = c(1, 2, 3, 4), 
                           labels = c("Low low", "Low", "Middle low", "Middle to high")))

summary(ensab$ses)
```

Income: M3_119

NMW: national minimum wage (salario minimo legal vigente)

```{r}
glimpse(ensab$M3_119)

#8 are NAs
ensab %>%
  filter(M3_119 == 8) %>%
  count()

#transform 7 unemployed into 0

ensab <- ensab %>%
  mutate(income = case_when(
    M3_119 == 8 ~ NA,
    M3_119 == 7 ~ 0,
    TRUE ~ M3_119)) %>%
  mutate(income = factor(income,
                         levels = c(0, 1, 2, 3, 4, 5, 6),
                         labels = c("Unemployed", "Less than half NMW", "Half - 1 NMW", "1-2 NMW", "2-3 NMW", "3-4 NMW", "More than 4 NMW")))

summary(ensab$income)

```

Health insurance coverage/scheme:

Chose REGIMEN over M3_113E because it has less categories plus no missing values, and it reflects the insurance status of the child and not the main provider.

```{r}
glimpse(ensab$REGIMEN)

#counting NAs in M3_113E

ensab %>%
  filter(M3_113E == 6) %>%
  count()

ensab <- ensab %>%
  mutate(insurance = case_when(
    REGIMEN == 4 ~ 0,
    TRUE ~ REGIMEN)) %>%
  mutate(insurance = factor(insurance, 
                           levels = c(0, 1, 2, 3), 
                           labels = c("Uninsured", "Contributory", "Subsidized", "Other")))

summary(ensab$insurance)
```

Region

```{r}
glimpse(ensab$Region)

ensab <- ensab %>%
  mutate(region = factor(Region,
                         levels = c(1, 2, 3, 4, 5, 6),
                         labels = c("Atlantic", "Eastern", "Central", "Pacific", "Bogota", "Orinoquia - Amazonia")))

summary(ensab$region)
```

Urban vs rural: Zona

Will be named location
Chose Zona over Clase because it has more universal categories (rural and urban), while Clase has categories according to the national census. However, it has more granularity and could be useful.

```{r}
glimpse(ensab$Zona)

ensab %>%
  group_by(Zona, .drop = FALSE) %>%
  count()

ensab <- ensab %>%
  mutate(location = factor(Zona,
                           levels = c(1, 2),
                           labels = c("Urban", "Rural")))

summary(ensab$location)
```

# Exploring the relationship between age of first appointment and caries (dmft)

The hypothesis is that the older the child was at it's first appointment, the higher the dmft index.

```{r}
ensab %>%
  filter(!is.na(agefdv)) %>%
ggplot(aes(x=agefdv, y=dmft)) +
  geom_boxplot() +
  facet_wrap(~age) +
  labs(x = "Age of first dental visit", y = "dmft Index score", title = "dmft Index score by age of first dental visit")
```

The expected results are seen in children 1 years old. However, in children 5 years old the children with the lowest median dmft are those in the "never" category. This can probably be confounded by the reason for the last dental appointment. 

It is probably better to explore the relationship by age category.

## Negative binomial regression

Subsetting the data stratified by age group: 

```{r}

#Subset the data for first dental visit

fdv <- ensab %>%
  dplyr::select(dmft, agefdv, age, sex, income, ses, insurance, region, location, ID_PERSONA, ID_HOGAR)

#by age group

fdv1 <- fdv %>%
  filter(age == 1)

fdv3 <- fdv %>%
  filter(age == 3)

fdv5 <- fdv %>%
  filter(age == 5)
```

Negative binomial models: crude and adjusted

MASS package used

```{r}
# Crude negative binomial regression age 1
model1 <- glm.nb(dmft ~ agefdv, data = fdv1)

summary(model1)
exp(model1$coefficients)
exp(confint(model1))
#theta is 0.03 so there is a lot of overdispersion

#Multiple negative binomial regression (adjusted model)
model1m <- glm.nb(data = fdv1, dmft ~ agefdv + sex + income + ses + insurance + region + location)

summary(model1m)
exp(model1m$coefficients)
#exp(confint(model1m)) is failing to converge

# Negative binomial regression age 3

model3 <- glm.nb(dmft ~ agefdv, data = fdv3)

summary(model3)
exp(model3$coefficients)
exp(confint(model3))

model3m <- glm.nb(data = fdv3, dmft ~ agefdv + sex + income + ses + insurance + region + location)

summary(model3m)
exp(model3m$coefficients)
exp(confint(model3m))

#Negative binomial regression age 5

model5 <- glm.nb(dmft ~ agefdv, data = fdv5)

summary(model5)
exp(model5$coefficients)
exp(confint(model5))

model5m <- glm.nb(data = fdv5, dmft ~ agefdv + sex + income + ses + insurance + region + location)

summary(model5m)
exp(model5m$coefficients)
exp(confint(model5m))

```

## Relationship between education on type of toothpaste and caries

```{r}
ensab %>%
  filter(!is.na(tp_info)) %>%
  ggplot(aes(x = tp_info, y = dmft)) +
  facet_wrap(~age) +
  geom_boxplot() +
  labs(x = "Information on toothpaste", y = "dmft Index score", title = "dmft Index score by information on toothpaste")

glimpse(ensab)
```

Negative binomial regression

```{r}
#Subset the data for toothpaste information

tp <- ensab %>%
  dplyr::select(dmft, tp_info, age, sex, income, ses, insurance, region, location, ID_PERSONA, ID_HOGAR)

modeltp1 <- glm.nb(data = tp, dmft ~ tp_info)
summary(modeltp1)

#ages 5

tp5 <- tp %>%
  filter(age == 5)

modeltp5 <- glm.nb(data = tp5, dmft ~ tp_info)
summary(modeltp5)
exp(modeltp5$coefficients)
exp(confint(modeltp5))
#not statistically significant

modeltp5m <- glm.nb(data = tp5, dmft ~ tp_info + sex + income + ses + insurance + region + location)
summary(modeltp5m)

#the relationship is attenuated after adjusting for confounders

#ages 3 and 5 

tp35 <- tp %>%
  filter(age == 5 | age == 3)

modeltp35 <- glm.nb(data = tp35, dmft ~ tp_info)
summary(modeltp35)
```

